A decentralized network support middleware for MMOGs

### 2009-09-22 trunk/textos/PlanoDoutorado/Lugano/planodoc_lugano.tex

This documente describes the full doctorate study plan whose part is intended to be executed in the University of Lugano, Switzerland. It proposes the research and creation of a novel support model for MMOGs (massively multiplayer online games), by using large-scale systems similar to those created for sharing of computing resources, such as file sharing peer-to-peer networks. Following this model, a distributed system could be formed to act as a server for an MMOG. Then, a portion of the game's communication and processing load would be delegated to each node of the server system -- more precisely, the game's virtual environment would be partitioned into regions, each of which managed by one or more server nodes. Such a system would be perceived by the players as an unique central servel to which they connect via a gateway. Upon connected, each player is seamlessly redirected to the appropriate server node. However, for such server system to work as expected, some issues must be addressed, such as maintaining the game state consistent among the different game server nodes, so each player has the same vision; management and balancing of the load assigned to the different server nodes; what to do when a node crashes or leaves the server system and, finally, how to broadcast and store the game state to the players in a distributed manner so that its most recent state can be restored even in the presence of system failures.

@@

In the past few decades, since the creation of the personal computer, computer games have enjoyed an evergrowing popularity. After the acess to the Internet became widely available to the general public, with increasingly fast and cheap domestic connections, multiplayer online games have become an important market. To illustrate this, a study from 2006 estimated that the online games market would grow from US\$ 3.4 billion in 2005 to US\$ 13 billion in 2011 \cite{intelligence2006online}. The main characteristic of the MMOGs is the large number of concurrent players interacting with one another through the Internet. This kind of game may be divided into several categories, such as:

@@

\item RPG: \emph{role-playing game}. In this kind of game, each player has a characters who evolves with time, acquiring power and accumulating wealth. These games are characterized by not having a predetermined end of the ``match'', creating a virtual world with persistent state. The player may, then, connect to this world, play for some time with his character, disconnect and, later, reconnect and continue from the point he was when he disconnected for the last time, for his state is stored in the game server.

@@

After single-player computer games evolved to multiplayer online games with the advent of the Internet, MMOGs emerged as a new trend because of the decrease of the domestic Internet connection cost and the increase of its average bandwidth. This kind of games have become very popular in the latest years because they allow the online interaction of a massive number of players at the same time. In the most succesful cases, such as World of Warcraft \cite{worldofwarcraft}, Lineage II \cite{lineage2} and EVE Online \cite{eveonline}, for example, tens of thousands of players are supported simultaneously \cite{chen2006pch}. These players can, then, interact with one another and with the virtual environment of the game.

@@

As already mentioned, MMOGs are characterized by a large number of players. To support this great number of participants, a client-server infrastructure is generally used, and all interaction is done through the server machine. In this paradigm, the server computer has the official copy of the game state and it is responsible for computing its change over time, while the game client is responsible for presenting to the user the game state sent by the server. The server is also responsible for updating the clients, in real time, on the changes that occur in the virtual environment of the game. Because of the high number of participants, the centralization on the game server makes the maintanance cost very high. Generally, it is required that the MMOG server has a connection to the Internet with several GBps of bandwidth and enough processing power to run the simulation of the virtual world \cite{feng2007wnn}. It becomes impossible, for example, the maintenance of massively multiplayer online games by small or independent groups on a limited budget.

@@

The main objective of this doctoral plan is to create a decentralized architectural model to support massively multiplayer online games. Following this model, it should be possible to form a system of geographically distributed server nodes, using the Internet as a means of connection between them. This distribution should be transparent to the players, to whom the system will be seen as a single MMOG server, which maintains the game world consistent, available and with a good quality of service. As a final product, a middleware for MMOG development with a high-level interface will be created.

@@

%TODO: mecanismo -> outra palavra
%TODO: dividir este item em 2: distribuiçao e rede overlay parecida com a real.
\item Develop a mechanism to deal with entry, exit and crash of server nodes, providing fault tolerance, while trying to keep the nodes which are geographically close to each other adjacent in the overlay network of the server system;

%TODO: dividir este item em 2
\item Create an algorithm for balancing the load of the game, maintaining the load on each server proportional to its capacity, while avoiding the splitting of a hotspot into different regions, as this would cause a strong dependence between these regions \cite{bezerra2009lbs};

\item Define a scheme for synchronizing the simulation state among different server nodes assigned to the same region of the gaming world, as well as synchronize periodically, or when necessary, the states in different regions, in order to maintain global consistency;

\item Build a distributed persistence mechanism to store the states of the players and of the virtual game world as a whole, so they are always available for any server node, even if some of these become unexpectedly unavailable;

\item Design techniques and algorithms that aim to reduce inter-server and server-client communication, in order to save resources of the server nodes, reduce delays in the communication between them and expand the audience of the game due to the reduction in the bandwidth required to play it and, besides, reducing the requirements for a computer to be part of the server system, without undermining the game experience of the players;

\item Create a middleware containing the desired services, along with a high-level API to make them accessible to the game developer.

@@

In a previous work \cite{bezerra2009lre}, it was set up a preliminary model -- which aimed to partition the virtual environment into regions, each assigned to a single server -- on which a load balancing scheme was defined. The Cosmmus -- Cooperative System for Massive Multiplayer Service -- is a new model to be defined during the doctorate course, which will also uses the principle of partitioning the virtual environment of the game into regions, but addressing the most critical problems (as referred in section \ref{researchquestions}), which were left out so far.

Cosmmus should include solutions to the major aspects of MMOGs, such as task distribution, synchronization of the simulation, distributed persistence and so on. Although this model is not defined yet, there are already some ideas on how it could be and what kind of approach it should use, based on related research. In the following sections it will be presented these initial ideas, which will be better developed during the doctoral course.
% Como os jogadores interagem!!!

@@

If a region get heavily crowded, one could think about splitting it in two or more new regions, but this could be counterproductive in certain situations, as in the case of a region that contains a hotspot: its division would result in two regions overly dependent on one another. Because of this kind of situation, it should be also created an algorithm, heuristic or metric to determine whether a region should or should not be subdivided into new regions.

It has been proposed, during the master course, a dynamic load balancing scheme based on graph partitioning and refinement of its partitions \cite{kernighan1970ehp, karypis1999pmk, karypis1998mam, dutt1993nfk}. The world was divided into cells of equal size and fixed position, each represented by a vertex in the graph. The interaction (measured in kilobytes per second) between the cells are represented by the edges of the vertices. The algorithm consists of partitioning the graph, where each partition represents a region to be assigned to a server. This balance takes into account different levels of resources among different servers and takes into account the needs of MMOGs, considering the bandwidth usage as the load to be balanced among the server nodes.

It will also be created a novel load balancing scheme which is intended to follow a few principles of the scheme proposed in the master's thesis. However, unlike the model suggested previously, there will be not only one server node per region, but a group of server nodes. In this case, it will be investigated ways to balance the load of the system, considering the distribution in two levels. Besides, other environment partitioning techniques could be used instead of using cell grids, such as space partitioning trees, as hinted in \cite{bezerra2009fgl}.

@@

Regarding the formation of the overlay network on which the server system will be deployed, there are some proposals, such as in \cite{ta2008nas, ratnasamy2002tao, han2005tao, zhang2004cla}, which try to build a network so that nodes with good network connection between them are adjacent in the overlay topology. It was proposed, for example, grouping the nodes according to the network delay between them. The goal is to use these techniques so that certain server nodes -- nodes which are serving adjacent regions and, especially, serving the same region -- belong to the same group. Adaptations are possible, including criteria other than delay, as the data transmission rate between them, in bits per second.

The purpose of mounting the overlay network following this strategy would be to minimize the delay in interaction between players. This delay in the interaction would be the time elapsed since the sending of an action from a player, processing by the server and receiving the resulting state by another player. If any two players are interacting with each other, but each one of them connected to a different server node, it will be required that each of their interactions is mediated by both servers. So, if the connection between these two servers has a low delay, the interaction of the players will be less harmed by the addition of a second intermediary.

@@

One of the key issues that will be handled by Cosmmus is maintaining the consistency of the simulation among the different server nodes. Even if each player sees just a part of the virtual world, this world must be the same for all servers. Thus, if two players are with their avatars in the same place of the virtual environment, they should see the same scene. Another important issue is the ordering of events. Assuming that two players are fighting in the game, the avatar of one of them shoots with a gun and the other avatar moves, the order of these events would be crucial to decide whether the latter player had his avatar dead, only wounded or completely avoided the bullet.

@@

Taking into consideration that there are several regions forming a contiguous world, and that in each one of them there is one or more servers, these issues to maintain the consistency of the simulation becomes even more difficult. It is expected that the division of the environment is seamless to the player. Then, when approaching a region boundary, the player should be able to see whatever is beyond it and to move freely across the border. For this, it will be necessary that the server to which he is connected to communicate with the one that serves the region beyond the border which is being viewed by the player. It will provide information such as the state of the environment and of avatars who are nearby and of other objects in the game.

In \cite{cecin2006pps}, it was proposed a state synchronization scheme that consists of two simulators: one conservative and one optimistic. The optimistic one processes each player action immediately after it is received and its resulting state becomes available to the player and something is show on his screen. The conservative simulator, however, divides the time in turns and only performs the actions of the players after performing a synchronization. At each turn, it waits for the arrival of the actions of all players involved -- or some explicit notification that they did nothing -- which ared then sorted and processed. Only after this step the conservative simulator advances to the next turn. In the case of the optimistic simulation deviate beyond a tolerable limit from the conservative one, the latter overwrites its state on the first, which continues from there. However, this scheme was proposed for a fully decentralized architecture, being necessary to analyze it and make necessary adjustments so it can be used in a server system distributed in two levels, as intended for Cosmmus. There are also other works that aim to maintain a consistent simulation, such as \cite{zhou2007cmh, moon2006emc, muller2006csm}, which will also be studied and possibly some of its principles will be part of the solution for consistency designed for the final model.

@@

Regarding the storage of the state of the virtual world and the players, it will be devised a way to do this in a distributed manner, since there will not be an unique central server. It might be simpler to choose one of the server nodes to save this information. However, this could cause the game state to depended on the availability and reliability of the chosen node and, even if these two conditions were satisfied, the node's connection could be easily saturated by the great number of requests, slowing down the recovery of the states stored in it.

Cosmmus should then use a distributed storage scheme, where portions of the state information are stored in different places. Furthermore, it should replication, so that the state becomes available even if some nodes of the system disconnect or crash unexpectedly. DHT-based overlay networks, such as Pastry \cite{rowstron2001psd}, Tapestry \cite{zhao2004trg}, Chord \cite{stoica2001csp} and Koord \cite{kaashoek2003ksd} seem to be possible solutions to provide the kind of distributed and redundant storage desired. It is intended, therefore, to investigate how suitable these networks really are for this purpose. There are simulators, such as OverSim \cite{baumgart2007ofo}, allowing the simulation of overlay networks, such as those mentioned, and through simulation it be will assessed the behavior of the system when using some of these logical networks.

@@

As the server nodes are, ideally, of low cost, it is necessary to save their resources as much as possible, but without harming the game quality. For example, it can be made some interest-based filtering, being sent to each player only what he can see, in order to save as much bandwidth between clients and servers, and between the servers. Following this same principle, servers from two different regions may have an old version of the state of each other, if there is no interaction between players whose avatars are in different regions. Thus, these servers do not need to update each other so often.

@@

Several works have been made in order to make filtering interest-based message filtering \cite{bezerra2008a3, ahmed2008dai, morgan2005imm, minson2005aim}. In addition to studying which one best fits the proposed model, and evaluate the possible adjustments that are necessary, other ways to save the resources of the server system will be investigated.

@@

In order to offer to application (game) developers an easy-to-use set of services, a high-level application programming interface must be created for Cosmmus. This set of services will compose the Cosmmus middleware, which will encapsulate all the low level operations regarding the distribution of the game. There are some options to define this distribution model -- and, also, the abstraction provided to the game developer --, like distributed shared memory, publish/subscribe, message passing and others. Each one of these presents both advantages and disadvantages, but the publish/subscribe seems to be the most suited for the development of MMOGs, since the entities (avatars and objects) present in the game's virtual world imply state update messages, in which every player may be interested or not.

@@

The fundamental idea behind Cosmmus is to form a server system on a peer-to-peer network, rather than indiscriminately distributing the simulation among all players, as are several proposals surveyed. With such separation, where only the server nodes could decide the progress of the game, it could be formed communities whose sole purpose would be to keep MMOGs up and running, organizing who would be allowed to participate in the system as a server node, with which permissions, and so on. There could be a scheme similar to that used in various free software projects, in Wikipedia \cite{adler2007cdr} and others, where the individual -- or company -- who had the initial idea gradually delegates authority to community members, according to how much each one cooperated, creating a hierarchy, with different levels of permission, but with all participating in a common effort.

@@

\chapter{Workplan and schedule}

This section describes the workplan, classes to attend and the intended schedule to perform the activities described in the previous sections. Section \ref{sec:metodologia} describes the activities to execute and the intermediate goals to fulfill the study plan; section \ref{sec:credits} lists the classes to attend and section \ref{sec:lugano} describes shortly the intended international cooperation in Lugano.

@@

One of the objectives of the doctoral course is to continue the research started during the Master's course, during which a technique for bandwidth usage optimization have been proposed: the A$^3$ algorithm \cite{bezerra2008a3}, which consists of a fine-tuned interest management, based on different relevance levels for each pair of entities in the game. Another work done during the Master's was the definition of a distributed server load balancing scheme \cite{bezerra2009lbs} which takes into consideration several points that have been considered critical: first, the distributed MMOG server system is not necessarilyh homogeneous, requiring the load balancing scheme to consider the different amounts of resources; second, the players may move their avatars to certain positions with a higher probability, leading to the formation of densely populated areas (hotspots) in the virtual environment; and, finally, the algorithms proposed tried to minimize the inter-server communication by using a graph partitioning approach.

>>>>

This document describes the full doctorate study plan whose part is intended to be executed in the University of Lugano, Switzerland. It proposes the research and creation of a novel support model for MMOGs (massively multiplayer online games), by using large-scale systems similar to those created for sharing of computing resources, such as file sharing peer-to-peer networks. Following this model, a distributed system could be formed to act as a server for an MMOG. Then, a portion of the game's communication and processing load would be delegated to each node of the server system -- more precisely, the game's virtual environment would be partitioned into regions, each of which managed by one or more server nodes. Such a system would be perceived by the players as an unique central server to which they connect via a gateway. Upon connected, each player is seamlessly redirected to the appropriate server node. However, for such server system to work as expected, some issues must be addressed, such as maintaining the game state consistent among the different game server nodes, so each player has the same vision; management and balancing of the load assigned to the different server nodes; what to do when a node crashes or leaves the server system and, finally, how to broadcast and store the game state to the players in a distributed manner so that its most recent state can be restored even in the presence of system failures.

@@

In the past few decades, since the creation of the personal computer, computer games have enjoyed an ever-growing popularity. After the access to the Internet became widely available to the general public, with increasingly fast and cheap domestic connections, multiplayer online games have become an important market. To illustrate this, a study from 2006 estimated that the online games market would grow from US\$ 3.4 billion in 2005 to US\$ 13 billion in 2011 \cite{intelligence2006online}. The main characteristic of the MMOGs is the large number of concurrent players interacting with one another through the Internet. This kind of game may be divided into several categories, such as:

@@

\item RPG: \emph{role-playing game}. In this kind of game, each player has a character who evolves with time, acquiring power and accumulating wealth. These games are characterized by not having a predetermined end of the ``match'', creating a virtual world with persistent state. The player may, then, connect to this world, play for some time with his character, disconnect and, later, reconnect and continue from the point he was when he disconnected for the last time, for his state is stored in the game server.

@@

After single-player computer games evolved to multiplayer online games with the advent of the Internet, MMOGs emerged as a new trend because of the decrease of the domestic Internet connection cost and the increase of its average bandwidth. This kind of games have become very popular in the latest years because they allow the online interaction of a massive number of players at the same time. In the most successful cases, such as World of Warcraft \cite{worldofwarcraft}, Lineage II \cite{lineage2} and EVE Online \cite{eveonline}, for example, tens of thousands of players are supported simultaneously \cite{chen2006pch}. These players can, then, interact with one another and with the virtual environment of the game.

@@

As already mentioned, MMOGs are characterized by a large number of players. To support this great number of participants, a client-server infrastructure is generally used, and all interaction is done through the server machine. In this paradigm, the server computer has the official copy of the game state and it is responsible for computing its change over time, while the game client is responsible for presenting to the user the game state sent by the server. The server is also responsible for updating the clients, in real time, on the changes that occur in the virtual environment of the game. Because of the high number of participants, the centralization on the game server makes the maintenance cost very high. Generally, it is required that the MMOG server has a connection to the Internet with several GBps of bandwidth and enough processing power to run the simulation of the virtual world \cite{feng2007wnn}. It becomes impossible, for example, the maintenance of massively multiplayer online games by small or independent groups on a limited budget.

@@

The main objective of this doctoral plan is to create a decentralized architectural model to support massively multiplayer online games. Following this model, it should be possible to form a system of geographically distributed server nodes, using the Internet as a means of connection between them. This distribution should be transparent to the players, to whom the system will be seen as a single MMOG server, which maintains the game world consistent, available and with a good quality of service. As a final product, a middleware for MMOG development with a high-level programming interface will be created.

@@

%TODO: mecanismo -> outra palavra
%TODO: dividir este item em 2: distribuiçao e rede overlay parecida com a real.
\item Develop a mechanism to deal with entry, exit and crash of server nodes, providing fault tolerance, while trying to keep adjacent, in the overlay network of the server system, the nodes which are geographically close to each other -- the purpose of this is to explore the good connection between them;

%TODO: dividir este item em 2
\item Create an algorithm for balancing the load of the game, maintaining the load on each server proportional to its capacity, while avoiding the splitting of hotspots into different regions, as this would cause a strong dependence between these regions; \cite{bezerra2009lbs};

\item Define a scheme for synchronizing the simulation state among the different server nodes, in order to maintain global consistency;

\item Build a distributed persistence mechanism to store the states of the players and of the virtual world as a whole, so that they are always available, even if some of the server nodes become unexpectedly unavailable;

\item Design techniques and algorithms to reduce inter-server and server-client communication, in order to save resources from the server nodes and reduce delays in the communication between them. As an additional result, the games will have a wider public -- due to the reduction of the bandwidth required to play -- and, also, the requirements for a computer to be part of the server system will be decreased;

\item Create a middleware containing the desired services, along with a high-level API to make them accessible and easy to use to the game developer.

@@

In a previous work \cite{bezerra2009lre}, it was set up a preliminary model -- which aimed to partition the virtual environment into regions, each assigned to a single server -- on which a load balancing scheme was defined. The Cosmmus -- Cooperative System for Massive Multiplayer Service -- is a new model to be defined during the doctorate course, which will also use the principle of partitioning the virtual environment of the game into regions, but addressing the most critical problems (as referred in section \ref{researchquestions}), which were left out so far.

Cosmmus will include solutions to the major aspects of MMOGs, such as task distribution, synchronization of the simulation, distributed persistence and so on. Although
this model is not yet defined, there are already some ideas on how it could be and what kind of approach it should use, based on related research. Upon finished, these ideas will be implemented as services in a middleware for developing MMOGs. The following sections describe these ideas, which will be better developed during the doctoral course.
% Como os jogadores interagem!!!

@@

If a region gets heavily crowded, one could think about splitting it in two or more new regions, but this could be counterproductive in certain situations, as in the case of a region that contains a hotspot: its division would result in two regions overly dependent on one another. Because of this kind of situation, it should be also created an algorithm, heuristic or metric to determine whether a region should or should not be subdivided into new regions.

It has been proposed, in previous works \cite{bezerra2009lbs, bezerra2009lre}, a dynamic load balancing scheme based on graph partitioning and refinement of its partitions \cite{kernighan1970ehp, karypis1999pmk, karypis1998mam, dutt1993nfk}. The world was divided into cells of equal size and fixed position, each represented by a vertex in the graph. The interaction (measured in kilobytes per second) between the cells are represented by the edges of the vertices. The algorithm consists of partitioning the graph, where each partition represents a region to be assigned to a server. This balancing allows for heterogeneous server systems -- that is, each node may have a different amount of available resources -- and takes into account the needs of MMOGs, considering the bandwidth usage as the load to be balanced among the server nodes.

It will also be created a novel load balancing scheme which is intended to follow a few principles of the scheme proposed in the previous works. However, unlike the model suggested previously, there will be not only one server node per region, but a group of server nodes. In this case, it will be investigated ways to balance the load of the system, considering the distribution in two levels. Besides, other environment partitioning techniques could be used instead of using cell grids, such as space partitioning trees, as hinted in \cite{bezerra2009fgl}.

@@

Regarding the formation of the overlay network on which the server system will be deployed, there are some proposals, such as in \cite{ta2008nas, ratnasamy2002tao, han2005tao, zhang2004cla}, which try to build a network so that nodes with good network connection between them are adjacent in the overlay topology. It was proposed, for example, grouping the nodes according to the network delay between them. The goal is to use these techniques so that certain server nodes -- nodes which are serving adjacent regions and, especially, serving the same region -- belong to the same group. Adaptations are possible, including criteria other than delay, such as the data transmission rate between them.

The purpose of building the overlay network following this strategy would be to minimize the delay in interaction between players -- this interaction delay is the time elapsed since the sending of an action from a player, processing by the server and receiving the resulting state by another player. If any two players are interacting with each other, but each one of them connected to a different server node, it will be required that each of their interactions is mediated by both servers. So, if the connection between these two servers has a low delay, the interaction of the players will be less harmed by the addition of a second intermediary.

@@

One of the key issues that will be handled by Cosmmus is maintaining the consistency of the simulation among the different server nodes. Even if each player sees just a part of the virtual world, this world must be the same for all servers. Thus, if two players are with their avatars in the same place of the virtual environment, they should see the same scene. Another important issue is the ordering of events. Assuming that two players are fighting in the game, and that the avatar of one of them shoots with a gun and the other avatar moves, the order of these events would be crucial to decide whether the latter player had his avatar dead, just wounded or completely avoided the bullet.

@@

Taking into consideration that there are many regions forming a contiguous world, and that in each one of them there is one or more servers, the virtual world simulation consistency problem becomes even more difficult to solve. It is expected that the division of the environment is seamless to the player. Then, when approaching a region boundary, the player should be able to see whatever is beyond it and to move freely across the border. For this, it will be necessary that the server to which he is connected to communicate with the one that serves the region beyond the border which is being viewed by the player. It will provide information such as the state of the environment and of avatars who are nearby and of other objects in the game.

In \cite{cecin2006pps}, it was proposed a state synchronization scheme that consists of two simulators: one conservative and one optimistic. The optimistic one processes each player action immediately after it is received and its resulting state becomes available to the player and something is show on his screen. The conservative simulator, however, divides the time in turns and only performs the actions of the players after executing a sort of synchronization. At each turn, it waits for the arrival of the actions of all players involved -- or some explicit notification that they did nothing -- which are then sorted and processed. Only after this step the conservative simulator advances to the next turn. In the case of the optimistic simulation deviate beyond a tolerable limit from the conservative one, the latter overwrites its state on the first, which continues from there. However, this scheme was proposed for a fully decentralized architecture, being necessary to analyze it and make necessary adjustments so it can be used in a server system distributed in two levels, as intended for Cosmmus. There are also other works that aim to maintain a consistent simulation, such as \cite{zhou2007cmh, moon2006emc, muller2006csm}, which will also be studied and possibly some of its principles will be part of the solution for consistency designed for the final model.

@@

Regarding the storage of the state of the virtual world and the players, it will be devised a way to do this in a distributed manner, since there will not be an unique central server. It might be simpler to choose one of the server nodes to save this information. However, this could cause the game state to depend on the availability and reliability of the chosen node and, even if these two conditions were satisfied, the node's connection could be easily saturated by the great number of requests, slowing down the recovery of the states stored in it.

Cosmmus should then use a distributed storage scheme, where portions of the state information are stored in different places. Furthermore, it should use replication, so that the state becomes available even if some nodes of the system disconnect or crash unexpectedly. DHT-based overlay networks, such as Pastry \cite{rowstron2001psd}, Tapestry \cite{zhao2004trg}, Chord \cite{stoica2001csp} and Koord \cite{kaashoek2003ksd} seem to be possible solutions to provide the kind of distributed and redundant storage desired. It is intended, therefore, to investigate how suitable these networks really are for the context of MMOGs. There are simulators, such as OverSim \cite{baumgart2007ofo}, allowing the simulation of overlay networks, such as those mentioned, and through simulation it be will assessed the behavior of the system when using some of these logical networks.

@@

As the server nodes are, ideally, of low cost, it is necessary to save their resources as much as possible, but without harming the game quality. For example, it can be made some interest-based filtering, being sent to each player only what he can see, in order to reduce the traffic between clients and servers, and between the servers. Following this same principle, servers from two different regions may have an old version of the state of each other, if there is no interaction between players whose avatars are in different regions. Thus, these servers do not need to update each other so often.

@@

Several works have been made in order to perform interest-based message filtering \cite{bezerra2008a3, ahmed2008dai, morgan2005imm, minson2005aim}. In addition to studying which one best fits the proposed model, and evaluate the possible adjustments that are necessary, other ways to save the resources of the server system will be investigated.

@@

In order to offer to application (game) developers an easy-to-use set of services, a high-level application programming interface must be created for Cosmmus. This set of services will compose the Cosmmus middleware, which will encapsulate all the low level operations regarding the distribution of the game.

There are some options to define this distribution model -- and, also, the abstraction provided to the game developer --, like distributed shared memory, publish/subscribe, message passing and others. Each one of these presents both advantages and disadvantages, but the publish/subscribe model seems to be the most suited for the development of MMOGs, since the entities (avatars and objects) present in the game's virtual world imply state update messages -- which may be published --, each of which may be relevant or not to each player -- who, therefore, may be subscribing or not to these state updates.

@@

The fundamental idea behind Cosmmus is to form a server system on a peer-to-peer network, rather than just distributing the simulation among all players, as are several proposals surveyed. With such separation, where only the server nodes could decide the progress of the game, it could be formed communities whose sole purpose would be to keep MMOGs up and running, organizing who would be allowed to participate in the system as a server node, with which permissions, and so on. There could be a scheme similar to that used in various free software projects, in Wikipedia \cite{adler2007cdr} and others, where the individual who had the initial idea gradually delegates authority to community members, according to how much each one cooperated, creating a hierarchy, with different permission levels, but with all participating in a common effort.

@@

\chapter{Work plan and schedule}

This section describes the work plan, classes to attend and the intended schedule to perform the activities described in the previous sections. Section \ref{sec:metodologia} describes the activities to execute and the intermediate goals to fulfill the study plan; section \ref{sec:credits} lists the classes to attend and section \ref{sec:lugano} describes shortly the intended international cooperation in Lugano.

@@

One of the objectives of the doctoral course is to continue the research started during the Master's course, during which a technique for bandwidth usage optimization have been proposed: the A$^3$ algorithm \cite{bezerra2008a3}, which consists of a fine-tuned interest management, based on different relevance levels for each pair of entities in the game. Another work done during the Master's was the definition of a distributed server load balancing scheme \cite{bezerra2009lbs} which takes into consideration several points that have been considered critical: first, the distributed MMOG server system is not necessarily homogeneous, requiring the load balancing scheme to consider the different amounts of resources; second, the players may move their avatars to certain positions with a higher probability, leading to the formation of densely populated areas (hotspots) in the virtual environment; and, finally, the algorithms proposed tried to minimize the inter-server communication by using a graph partitioning approach.
